
\chapter{Networks and the Internet}
\label{chap:networks-internet}

\chapterintro{Modern LLM applications usually communicate over networks. This chapter introduces clients, servers, requests, responses, and APIs.}
\learningobjectives{\begin{itemize}\item Explain what a network is.\item Describe clients and servers.\item Understand HTTP at a beginner level.\item Connect APIs to LLM applications.\end{itemize}}

\section{What Is a Network?}
A network is a collection of connected computers that can exchange information.

\section{Clients and Servers}
A client asks for something. A server responds. When you use a chatbot in a web browser, the browser acts as a client and the AI service acts as a server.

\section{HTTP}
HTTP is a common protocol used by web applications. A request asks for a resource or action. A response returns information.

\section{APIs}
An Application Programming Interface defines how software components communicate. LLM applications often expose APIs so other programs can send prompts and receive generated responses.

\section{Python Example}
\begin{lstlisting}
# This example uses only the standard library.
from urllib.request import urlopen

with urlopen("https://example.com") as response:
    html = response.read(100)
    print(html)
\end{lstlisting}

\section{Exercise}
\begin{exercise}Describe the client and server in an AI chatbot interaction.\end{exercise}
\begin{solution}The user's browser or app is the client. The machine running the chatbot backend and model service is the server.\end{solution}

\section{Summary}
Networks allow AI systems to serve users. Understanding requests, responses, and APIs prepares the learner for production LLM engineering.
