\chapter{What Is a Computer?}
\label{chap:what-is-a-computer}
\chapterintro{This chapter introduces the basic idea of a computer and connects it to AI engineering.}
\learningobjectives{\begin{itemize}\item Define what a computer is.\item Explain input, processing, storage, and output.\item Distinguish hardware from software.\item Understand why these ideas matter for LLM systems.\end{itemize}}
\section{Introduction}
Before studying Python, Databricks, machine learning, transformers, or large language models, we need to answer a simple question: what is a computer?
\begin{definition}A computer is an electronic programmable machine that receives input, processes data according to instructions, stores information, and produces output.\end{definition}
\section{The Four Basic Operations}
Every computer performs four operations: input, processing, storage, and output.
\subsection{Input}Input is information entering the computer. Examples include a keyboard, microphone, camera, file, or network request.
\subsection{Processing}Processing means applying instructions to data. CPUs handle general processing. GPUs are often used for deep learning because they can perform many numerical operations in parallel.
\subsection{Storage}Storage lets computers remember information. RAM stores temporary working data. SSDs and hard drives store longer-term data.
\subsection{Output}Output is information produced by the computer, such as text on a screen, a saved file, or a response from an AI chatbot.
\section{Hardware and Software}
Hardware is physical equipment. Software is instructions. An LLM application is software running on hardware.
\section{A First Python Example}
\begin{lstlisting}
print("Hello, LLM Engineer!")
\end{lstlisting}
\gitcheckpoint{After creating this repository, commit the initial LaTeX structure with \code{git add .} and \code{git commit -m "Initial book structure"}.}
\section{Exercises}
\begin{exercise}Classify keyboard, CPU, SSD, monitor, and microphone as input, processing, storage, or output.\end{exercise}
\begin{solution}Keyboard and microphone are input. CPU is processing. SSD is storage. Monitor is output.\end{solution}
\section{Summary}A computer receives input, processes data, stores information, and produces output. These ideas support all later AI engineering topics.
